\section{Implementation}
\label{sec:implementation}

Figure~\ref{fig:architecture} shows the system as deployed during the
tournament: a collector for the official FIFA Fantasy API, external
fetchers for match history \citep{martj42} and club-level data
\citep{footballdata}, a feature builder producing one table row per
player and round, the three prediction backends of
Section~\ref{sec:methodology} with their per-position ensemble, a
mixed-integer squad optimizer, and a live decision layer for captain
and substitution calls. This section describes the pipeline
architecture, the engineering decisions that mattered in practice,
our reproducibility and testing discipline, and the team-news
ingestion subsystem.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
      node distance=6mm and 9mm,
      box/.style={draw, rounded corners=1.5pt, align=center,
                  minimum height=8mm, inner sep=2.5mm, font=\small},
      ext/.style={box, dashed},
      arr/.style={-{Stealth[length=2.2mm]}, semithick},
    ]
    \node[box] (collector) {FIFA Fantasy API\\collector};
    \node[box, right=of collector] (features) {Feature\\builder};
    \node[box, right=of features] (models) {Backends: heuristic,\\Poisson, GBM};
    \node[box, below=of models] (ensemble) {Per-position\\ensemble};
    \node[box, left=of ensemble] (optimizer) {MILP squad\\optimizer};
    \node[box, left=of optimizer] (live) {Live decision\\layer};
    \node[ext, below=of collector] (elo) {International Elo\\(match history)};
    \node[ext, below=of features] (news) {Team news and\\news signals};
    \draw[arr] (collector) -- (features);
    \draw[arr] (features) -- (models);
    \draw[arr] (models) -- (ensemble);
    \draw[arr] (ensemble) -- (optimizer);
    \draw[arr] (optimizer) -- (live);
    \draw[arr] (elo) -- (features);
    \draw[arr] (news) -- (features);
  \end{tikzpicture}
  \caption{System architecture: data collection, feature construction,
    the three prediction backends and their ensemble, squad
    optimization, and the live decision layer. Dashed boxes are
    external signal sources.}
  \label{fig:architecture}
\end{figure}

\subsection{Pipeline architecture}
\label{sec:implementation-pipeline}

The pipeline is a linear chain of stages mediated entirely by files
on disk. The collector snapshots the three FIFA Fantasy endpoints
into raw Parquet tables; the external module fetches international
match history and club-level CSVs and derives the country and club
Elo tables; the feature builder joins these into a per-player,
per-round feature table; each model backend reads that table and
writes a predictions table; the optimizer reads predictions and
writes a recommendation as JSON and Markdown; and a report generator
renders static HTML from the accumulated JSON. Because every stage
reads from and writes to disk, the pipeline is idempotent: any stage
can be re-run with different parameters without re-running its
upstream stages, and only the collector and external fetchers touch
the network at refresh time.

The repository mirrors this dataflow. The Python package under
\texttt{src/fifa\_fantasy/} holds one subpackage per stage
(\texttt{collector}, \texttt{external}, \texttt{features},
\texttt{model}, \texttt{optimizer}, \texttt{live}, \texttt{web}),
plus a single canonical scoring module shared by training,
validation, and reporting. Data directories separate raw collector
snapshots, derived external tables, processed feature and prediction
tables, training dumps, and per-run recommendation outputs; the test
suite and ad-hoc analysis scripts live alongside. Model artefacts
are stored in LightGBM's text format so they diff cleanly under
version control.

\subsection{Engineering decisions}
\label{sec:implementation-engineering}

Four decisions deserve explicit mention.

\paragraph{Parquet as the interchange format.}
Every intermediate table is Parquet rather than CSV: typed columns,
smaller files, and faster reads across the many re-runs a live
tournament demands. The only hand-maintained CSV is the static FIFA
ranking snapshot of Section~\ref{sec:data-rankings}.

\paragraph{Self-documenting result files.}
Every recommendation file is named with the originating hostname,
backend, stage, and a UTC timestamp. This prevents two collaborators
from clobbering each other's runs and makes the results directory an
audit trail: the JSON carries the full squad with per-player
metadata, lineup and captain decisions, transfer blocks, and
quantile bands where the GBM backend supplies them.

\paragraph{Deterministic training.}
All LightGBM \citep{ke2017lightgbm} training runs pin the seed, the
bagging and feature-fraction seeds, and the library's deterministic
flag. Without this, held-out RMSE wobbles from run to run by enough
to mask small feature-ablation effects; determinism was a
precondition for the A/B comparisons of
Section~\ref{sec:negative-results}.

\paragraph{Exact optimization over greedy selection.}
Squad selection under a budget cap, per-country limits, position
quotas, and a transfer quota has too many interacting constraints
for a greedy heuristic to be trustworthy. We formulate it as a
binary integer program in PuLP \citep{pulp} and solve with CBC
\citep{cbc}; the instance solves in under a second on commodity
hardware, so exactness costs nothing operationally.

\subsection{Reproducibility}
\label{sec:implementation-reproducibility}

Every direct dependency is pinned to a known-good version, model
training is deterministic given the seed, and Monte Carlo
simulations use a fixed NumPy generator seed. The static inputs (the
ranking snapshot, the cached match-history CSV, the club-level CSVs)
are committed to the repository or cached under a versioned
directory, and the pre-tournament FIFA Fantasy snapshot is committed
as well, so a fresh clone reproduces the full pipeline end-to-end
without any network access.

\subsection{Testing}
\label{sec:implementation-testing}

The pytest suite comprises \TestCount{} tests at the pre-final
freeze. The largest share pins the scoring functions:
parametrised cases cover every per-position scoring component
(appearance, goals, assists, clean sheets, conceded-goal penalties,
save bonuses, and the ownership-threshold conditions of the scouting
bonus), together with realistic end-to-end scenarios pinned to known
point totals. The collector, feature builder, and optimizer each
have dedicated test modules backed by end-to-end pipeline runs.
Model correctness is guarded by the held-out validation on the
English Premier League data of Section~\ref{sec:data-fpl}
\citep{vaastav}: any backend change that drifts held-out RMSE by
more than 0.05 in either direction is caught when the validation is
re-run.

\subsection{Team-news ingestion}
\label{sec:implementation-team-news}

Expert fantasy practitioners read manager press conferences roughly
24 hours before kickoff to anticipate rotation, suspensions, and
lineup changes. Our pipeline initially had no such input, so we
built a team-news subsystem that converts published pre-match
previews into a per-fixture predicted-XI signal.

\paragraph{The predicted-XI signal.}
Per-source parsers extract raw lineups (player names as printed,
with a starting or bench status) from public previews and persist
them as one row per fixture, player, and status. The feature builder
joins the most recent lineup per fixture into the feature table as
two columns: a three-valued starting indicator (confirmed start,
confirmed bench, or missing when no news exists) and a confidence
score in $[0,1]$ inherited from the source. The join is NaN-safe by
construction: with no news, both columns are missing and every
backend behaves exactly as before, which we verified by re-running
the backtest with the wiring in place and no data (all backend
totals were unchanged).

\paragraph{Name matching.}
Scraped names must be resolved to FIFA player identifiers. The
matcher normalises case and strips accents, then applies tiered
rules with descending confidence: an exact full-name match (1.00), a
match on the display name such as a mononym (0.95), a last-name
match within the player's country and position (0.85), within the
country alone (0.70), and a last-name match with no country context
(0.50). Ambiguous candidates at any tier abort the match rather than
guess, and unmatched names are logged for manual review; multi-part
Latin American and African naming conventions are the main source of
false negatives.

\paragraph{Availability discount.}
The lineup signal feeds availability handling in two ways. First, a
hard gate: the deterministic backends zero the prediction of any
player with a confirmed bench status, exactly as they do for
transferred or eliminated players, while the Monte Carlo backend
scales simulated points by the confidence score instead of zeroing,
damping low-confidence reports. Second, a soft participation
discount applied to effective points just before optimization: each
player's recent participation rate (the share of their team's
completed matches in which they took the pitch) maps linearly to a
multiplier between a floor of 0.5 and 1. The floor is deliberately
non-zero because participation is measured by a scored-points proxy,
so a nailed starter who blanked twice would otherwise be
misclassified as a rotation risk. Missing participation history
means factor 1 before the tournament (no claim either way), but once
any player has history, a missing value means the player has never
taken the pitch and is treated as the floor; without this rule the
optimizer favoured cheap third-choice goalkeepers whose team-level
clean-sheet points equalled the starter's.

\paragraph{What actually produced data.}
We implemented three source paths: an integration with a
community-maintained scraping library (scaffolded pending confirmed
World Cup coverage), a parser for ESPN pre-match preview articles
that requires the operator to supply article URLs, and a third-party
predicted-lineup site whose parser was never tuned. In practice only
the manual ESPN path produced predicted-XI rows: the scheduled
unattended runs wrote empty snapshots throughout the tournament, so
the signal reached the models only when we hand-fed article URLs
ahead of selected fixtures, with coverage skewed toward high-profile
matches. A companion collector for coarser news signals (headline
and summary feeds filtered by tournament keywords) supplies the
second dashed input of Figure~\ref{fig:architecture}. The scraping
engineering underneath both (client fingerprinting, caching, rate
limiting, retries) is deliberately generic and separated from the
fantasy-specific parsers; we document it in the repository rather
than here.
