\section{Conclusion}
\label{sec:conclusion}

% Final numbers land after the 2026-07-19 final; the two remaining
% placeholders are the only edits this section still needs.

We set out to answer a practical question: how much of a fantasy
football tournament can be handled by models that are built, validated,
and revised while the tournament is running? The system that emerged --
three complementary point-prediction backends routed per position by
held-out error, a mixed-integer squad and transfer optimizer that now
prices captaincy ceiling, availability, and the backup-goalkeeper
asymmetry, and a live decision layer audited round by round against its
own recommendations -- answered most of it. Across the seven rounds
completed before the final, every backend beat a constraint-valid
random baseline by a wide margin (\TotalGbm{} points for the strongest
backend and \TotalEnsemble{} for the production ensemble against
\TotalRandomBaselineMean{} for chance), and the human-plus-model entry
scored \TotalUserActualNet{} net points while paying real transfer
costs the replays do not. The record closed with the entry's best
round of the tournament, \FinalRoundNet{} points on the locked
final-round plan, lifting it to global rank \GlobalRankFinal{}
among the game's worldwide entries.

The negative results are, we believe, the most transferable part of
the record. A six-component upgrade lost to the baseline it meant to
replace; a theoretically derived goalkeeper multiplier measured worse
than an empirically calibrated one at less than half its size;
prediction-market prices failed to improve player-level forecasts as a
regression feature, earned only a capped minority weight in match-level
forecasting, and lost to a plain Elo rating out of sample; and minutes
information, whether proxied or real, helped only in the selection
layer, never in the point model. The positive findings mirror them:
leak-free trailing form, tournament retraining, real expected-goals
form, per-position routing, and ceiling-aware captaincy each cleared a
pre-registered walk-forward gate before deployment. None of these
lessons is specific to this World Cup; they are lessons about small
samples, about anchoring on sentiment, and about the discipline of
validating every change on data the model has never seen.

The final round is a fitting closing exhibit for the collaboration
this paper documents, because credit divides cleanly. The model
supplied the structure: the transfer plan into the eventual
champion's defensive block, the booster selection, and the captain
whose doubled hat-trick was the entry's largest single return. The
operator supplied the corrections the model could not see: the
insight that only one goalkeeper can score per round (adopted into
the optimizer's objective the same day), the rotation read that
benched a forward who never played, and the discovery that the app
locks the captaincy at kickoff, which forced a static commitment
the tooling had modelled as flexible. The model's one clear miss --
a ten-point scorer left on the bench of the model-guided eleven --
belongs in the ledger next to the overrides it got right, as does
the season-long pattern that the entry's best rounds were the ones
where model and operator agreed. The complete system, the frozen
data snapshots behind every table, and the round-by-round decision
log with realized outcomes are released as open source, so that the
next attempt at this problem can start where this one finished
rather than where it began.
