\section*{Author Contributions}

This work combines computational and football-domain expertise. The
contributions are stated explicitly to distinguish decisions derived
from data and code from those derived from match-watching judgment.

\textbf{Asher Davila} (computational lead): conceptualization, system
architecture, and software, including the three predictor backends,
the MILP optimizer, the live decision tools, and the data ingestion
pipeline; all held-out validation and hyperparameter sweeps; writing
of the code and of this manuscript; and project administration. He
operated the fantasy entry through every matchday and made the final
call on every transfer, captain choice, and formation.

\textbf{Diego Guajardo} (domain expert): football domain knowledge
from direct observation of the qualifying rounds, the pre-tournament
friendlies, and the tournament matches. He supplied per-player
judgments that the predictive models could not derive from features
alone: minutes risk inferred from observed manager behavior, in-form
versus out-of-form distinctions between players with similar predicted
scores, and tournament-narrative signals such as top-scorer race
incentives and must-win match dynamics. He reviewed the squad through
every transfer round, applying match-observation context to the model
outputs before each final pick.

Specific calls credited to D.~Guajardo include the pre-tournament
inclusion of Willian Pacho in the matchday-one squad, a low-ownership
defender who returned strongly relative to price; the rotation-risk
flags identifying which clinched-group teams would rest starters late
in the group stage; and the form read that decided the matchday-two
captain switch from Lautaro to Olise, which produced the squad's
largest captain return.

\begin{center}
\begin{tabular}{lll}
\toprule
Activity & A.~Davila & D.~Guajardo \\
\midrule
System architecture and software & lead & reviewing \\
Model development and validation & lead & reviewing \\
Football domain knowledge & supporting & lead \\
Transfer and captain decisions & shared & shared \\
Pre-tournament squad construction & shared & shared \\
Writing & lead & reviewing \\
Project administration & lead & none \\
\bottomrule
\end{tabular}
\end{center}

The two skill sets are complementary: the decisions that performed
well combined both signals, while the decisions that performed poorly
weighted one signal over the other unnecessarily.
